Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-09-06 analysis/paper/paper.Rmd
Minor edits following comments from David

# Introduction
Variation in population patterns are regarded as important drivers and potential outcomes of cultural variation, and is thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates have become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].

The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @solheim2018; @fossum2020; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @crema2022 for a recent review].

The most critical of these follow from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem cannot be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made. However, it is worth underscoring that deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5th and 97.5th quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not take into account that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

When the frequency distribution of shoreline dates have been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline dated sites, where the probability of drawing any 5 year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *ADMUR*.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically [@crema2022; @timpson2021], found by using the values of the final SPD. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values. The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs some time around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Subsequently, the SSPD deviates negatively from the exponential null from c. 6800 to c. 5800 BCE. This is followed by a positive deviation in the last centuries before 5000 BCE. These developments should possibly been seen in relation to simultaneous developments in the inland areas of southern Norway. The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. @boaz1999 related this to an economic reorganisation, involving a more focused exploitation of the interior and a reduction in the utilisation of coastal areas. This period is followed by a marked reduction in inland activity, lasting until c. 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c. 5800 BCE and the positive deviation from c. 5300--5000 BCE. @boaz1999 associated this period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

@@

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.

>>>>

# Introduction
Variations in population patterns are regarded as important drivers and potential outcomes of cultural variation, and are thus of critical importance to our understanding of past human societies [e.g. @french2021; @shennan2000; @shennan2021]. During the last decade, the frequency distribution of radiocarbon dates has become the preferred proxy for archaeologists studying prehistoric population dynamics [@crema2022; @french2021; @shennan2013], owing in part to the fact that radiocarbon dates represent the most temporally and spatially widespread archaeological data available for studying past population densities [@bird2022]. This has also been the case for Norwegian Stone Age archaeology [@solheim2018; @solheim2020a; @nielsen2019; @nielsen2021; @bergsvik2021; @jorgensen2020; @lundstrom2023; see also @persson2009].

The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987]. In this paper, we report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

@@

Previously, the frequency of shoreline dated sites has been evaluated by finding point estimates of shoreline dates that are aggregated in somewhat arbitrary bins of 200 or 500 years [e.g. @fossum2020; @solheim2018; @mjærum2022; @tallavaara2020], or by the use of disjoint time-intervals of variable duration [@berg-hansen2022], which could cause issues when comparing the relative counts across these units [@bevan2021]. However, the point-based estimates do not consider the uncertainty in the distance between the sites and the contemporaneous shoreline, nor the impact the variability in the rate of sea-level change has on the precision of the dates that can be achieved with the method. In a recent study, @roalkvam2023 has presented a probabilistic method for shoreline dating that takes these parameters into account, thus setting the stage for a more refined investigation of the relationship between the frequency of ^14^C-dates and shoreline dated sites. The parametrisation of the method was based on simulating the distance between 66 sites with ^14^C-dates and the prehistoric shoreline along the Skagerrak coast in the region between Horten municipality in the north-east to Arendal municipality in the south-west (see map in Figure \@ref(fig:map)). The results of the analysis indicates that the sites tend to have been located close to the shoreline until just after 4000 BCE when a few sites become more withdrawn from the shoreline [see also @nielsen2021a], followed by a clear break around 2500 BCE, at which point shoreline dating appears to lose its utility. Thus, this geographical and temporal limit are also used for this study.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently by factors such as those mentioned above. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @crema2022 for a recent review].

The most critical of these follows from the fact that the summation of the probabilities associated with the dates for SPDs is not a directly coherent statistical procedure. This is because the summed probabilities can no longer been seen as probabilities, but rather represent the combination of events and chronological uncertainties, making the two indistinguishable, and rendering the interpretation of the resulting sum difficult [@blackwell2003; @crema2022]. As @timpson2021[2] put it: 'the SPD is not *the* single best explanation of the data, nor even *a* single explanation of the data, but rather a conflation of many possible explanations simultaneously, each of which is mired by the artefacts inherited from the calibration wiggles.' The SPD is not a model. It is the combined representation of a range of possible explanations for the data -- the frequency of dated events combined with the variable uncertainty associated with these [@carleton2021]. This means that a direct visual inspection of SPDs can be misleading, that they cannot be directly analysed to draw inferences on population dynamics, and that they cannot be directly compared to other time-series data [@carleton2018; @timpson2021; @crema2022]. While this problem cannot be entirely resolved, a range of approaches have been developed in an attempt to work around this issue.

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE). The result from these simulations is then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, for example, an exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether the null model as a whole can be rejected as an explanation of the data, and it does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and it is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made. However, it is worth underscoring that deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

The final RSPD was subjected to the standard null-hypothesis testing approach through Monte Carlo simulation, as introduced above [see @shennan2013; @timpson2014], by fitting an exponential, logistic and uniform model to the observed RSPD. Following @timpson2021, the model parameters were identified by MLE search with *ADMUR*, using the differential evolution optimization algorithm *DEoptimR* [@brest2006]. The Monte Carlo simulations were then performed using the *SPDsimulationTest()* function from *ADMUR*. For each model, a series of individual calendar years are drawn from the model distribution, with replacement, the number of which equals the number of bins in the observed RSPD. These are then 'uncalibrated' to a single value on the ^14^C scale and a random error from among the observed errors is added to the date. These are then calibrated back to the calendar scale and finally summed. Here, this procedure was repeated 10000 times for each null model. The 2.5 and 97.5 quantile of the resulting summed probability for each year across all simulations are then retrieved to create the 95% critical envelope with which to compare the observed RSPD. The degree to which the observed RSPD deviates from the simulations is used to calculate a global p-value, indicating whether the null model under consideration can be rejected.

@@

Four geological reconstructions of shoreline displacement in the region lays the foundation for the method as implemented here (Figure \@ref(fig:shoredate)B). These shoreline displacement curves are from Horten [@romundset2021], Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b] and Arendal [@romundset2018b], each associated with a shoreline isobase along which the trajectory for relative sea-level change has been the same [see @svendsen1987; and map in Figure \@ref(fig:shoredate)A]. The first step in the dating procedure is to interpolate the shoreline displacement to the site to be dated. This is done by inverse distance weighting [e.g. @conolly2020], where the relative sea-level change is interpolated to the location of an archaeological site based on its distance from the isobases of the geological displacement curves.

@@

With an estimated trajectory of shoreline displacement at the site to be dated, the elevation of the site can be drawn on to find a likely date for when it was in use, under the assumption that it was in use when located on or close to the shoreline. Figure \@ref(fig:shoredate)C indicates where the mean elevation of the polygon representing the example site intersects the interpolated displacement curve. As it can reasonably be assumed that a site was not in use when located under water, this effectively represents a *terminus post quem* (TPQ) date. However, TPQ dates that do not account for the likely distance between site and sea when the site was in use limits the further inferential steps that can be taken. Furthermore, the common practice of adding a constant error or binning point-estimates of shoreline dates, as mentioned above, does not consider that the rate of shoreline displacement varies both spatially and temporally, and therefore introduces bias to the result on both the spatial and temporal scale, while also overestimating the precision of the method in most cases.

@@

When the frequency distribution of shoreline dates has been investigated in the past, this has typically involved finding point-estimates of dates by either using the mean of the displacement curves, or by taking mean of the TPQ dates for each site (as illustrated in Figure \@ref(fig:shoredate)C). As outlined above,  these point-estimates are then typically aggregated in bins of 200 or 500 years. This approach is illustrated in Figure \@ref(fig:sumshore)B. The SSPD resulting from using the probabilistic method of shoreline dating, given in Figure \@ref(fig:sumshore)C, demonstrates that accounting for the site-sea relationship gives a substantially different result compared to the more traditional approach, which is indicated by the shifts in the positions of peaks and troughs in the distribution after c. 8000 BCE.

@@

The Monte Carlo simulations with the SSPD is based on drawing a sample of calendar dates from the observed date range in the SSPD, equalling the number of shoreline-dated sites, where the probability of drawing any 5-year interval is determined by the null model of choice. This is equivalent to the sampling method *calsample* from *rcarbon* [@crema2021]. Each sampled date is then assigned one of the pre-interpolated displacement curves, the probability of which is weighted by the density of observed sites within each 2 km interval. The calendar date is 'uncalibrated' -- to follow the RSPD terminology -- to an elevation range from which a single elevation value is drawn with uniform probability between the lower and upper limit of the displacement curve at that calendar year, using intervals of 5 cm. As shoreline dating is here done with a gamma distribution with the same parameters across all sites, there is no equivalent of the error term for ^14^C-dates that determine the shape of the Gaussian distribution associated with the ^14^C age. Consequently, the elevation value retrieved by 'uncalibrating' the shoreline date was shoreline dated using the displacement curve for the relevant 2 km interval with the same gamma distribution for all samples. Having dated the number of samples equalling the number of shoreline-dated sites, these were then summed, and the entire process repeated a total of 10000 times. The 97.5% highest and 2.5% lowest summed probability at each interval of 5 years across all simulations was then retrieved to create the 95% critical envelope, and the global p-value found in the same manner as for the RSPDs. This was done using re-purposed code from *ADMUR*.

@@

For each dating method, the relative performance of the exponential, logistic and uniform models was compared. This was done by finding the overall relative log-likelihoods of the models using the *loglik* function from *ADMUR*. For any given model, this involves finding the scalar product between the probability of each shoreline date against the probability of the null model, and in the case of the ^14^C-dates, each site-phase. The overall model likelihood is the product of these individual probabilities and is not, critically, found by using the values of the final SPD [@crema2022; @timpson2021]. From these likelihoods, the Bayesian (or Schwarz) information criterion (BIC) was then found for each model. This penalises the models for the number of parameters in use to avoid over-fitting, and provides a measure for the relative performance of the models.

@@

```{r spdmc, echo = FALSE, fig.cap = "Monte Carlo simulation for shoreline and radiocarbon dates. A) Summed probability of shoreline-dated sites (n = 921) compared to an exponential null model. B) SSPD compared to a logistic null model. C) SSPD compared to a uniform null model. D) Summed probability of calibrated radiocarbon dates (n = 310, bins = 134) compared to an exponential null model. E) RSPD compared to a logistic null model. F) RSPD compared to a uniform null model.", out.width = "500px", fig.align="center", regfloat = TRUE}
knitr::include_graphics(here("analysis/figures/mc.png"))
```
The results of the Monte Carlo simulations are given in Figure \@ref(fig:spdmc), displaying the deviations from the simulation envelopes and the corresponding global p-values. The comparison of the models based on BIC scores is given in Figure \@ref(fig:bic).

@@

All of the standard models could be rejected for the SSPD. Of the contrasted models, the BIC values indicate that the exponential model is the best candidate. Its negative direction indicates that some overall process of reduction in the frequency of shoreline-dated sites occurs over time, while the rejection of the model and clear deviations from the exponential pattern indicates that this model does not account for the entire development (Figure \@ref(fig:spdmc)A). Some significant deviations occurs with the model over-predicting the frequency of sites in the period from c. 9200--9000 BCE, followed by a significant positive peak just after 9000 BCE, lasting until around 8500 BCE. The process of decline is then interrupted by positive deviations at 7900--7400, and with a smaller deviation around 7000 BCE. This is closely followed by a negative deviation between c. 6800 and 5800 BCE. Finally, a positive deviation occurs again around 5300--5000 BCE. While all of the above deviations should also be viewed with caution, questions concerning the reliability of shoreline dating after 4000 BCE [cf. @roalkvam2023] and the small magnitude of the negative deviations that occur around 3000 BCE means that these are disregarded in the discussion to follow.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs sometime around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

Subsequently, the SSPD deviates negatively from the exponential null model from c. 6800 to c. 5800 BCE. This is followed by a positive deviation in the last centuries before 5000 BCE. These developments should possibly be seen in relation to simultaneous developments in the inland areas of southern Norway. The negative deviation in the SSPD appears to correspond to an increase in inland activity, which is to take place in the period from c. 6700--5900 BCE and reach its peak around 6200--6000 BCE [@boaz1999; @selsing2010; @persson2018]. @boaz1999 related this to an economic reorganisation, involving a more focussed exploitation of the interior and a reduction in the utilisation of coastal areas. This period is followed by a marked reduction in inland activity, lasting until c. 4500--4000 BCE [@persson2018, 204], the onset of which thus corresponds to the end of the negative deviation in the SSPD at c. 5800 BCE and the positive deviation from c. 5300--5000 BCE. @boaz1999 associated this period of decrease in inland activity to the onset of the classic Nøstvet phase, which is now believed to commence in the middle of the 6th millennium BCE [@reitan2022]. The Nøstvet period is argued to be characterised by an increase in specialised coastal adaptation that also coincides with what @fuglestvedt2018[17] has been termed \'the great wave\' of rock art in Norway, starting from around 5700 BCE in south-eastern Norway [@fuglestvedt2018, 42--46; @glorstad2010, 216--233]. The onset of the classic Nøstvet thus roughly corresponds with the positive deviation in the SSPD around the later parts of the 6th millennium BCE.

@@

Another relevant point to mention here is that the relative sea-level change is effectively built into the Monte Carlo simulations, much in the same way as the Monte Carlo simulation method used with radiocarbon dates account for fluctuations in the calibration curve [e.g. @crema2022]. The simulation envelopes therefore pertain to the frequency of shoreline dated sites that could be expected under the null models, given the distribution of sites in the study area (Figure \@ref(fig:incpolys)A), the corresponding local trajectories of relative sea-level change (Figure \@ref(fig:incpolys)B), and a constant relationship between sites and shoreline described by the gamma function employed for the dating procedure (Figure \@ref(fig:shoredate)D). Beyond uncertainties to do with the dating method itself, and by extension also the generation of the Monte Carlo envelope, it is also conceivable that variation in rates of sea-level change has impacted settlement patterns beyond that which could be directly expected under these null models. An increased movement instantiated by increased rates of relative sea-level change could result in the transition to a settlement pattern involving more frequent moves than that which could be expected under a constant adjustment to the shoreline. If this is the case, the proxy would still reflect the degree of occupational duration, but this could be a response to the rate of relative sea-level change rather than other cultural and societal dimensions such as migratory events or adjustments to other environmental developments. Directly accounting for variation in the rate of sea-level change could thus prove important for properly assessing the substantive implications of the fluctuations in this proxy.
